---
output: 
  pdf_document:
      number_sections: true
      citation_package: natbib 
fontsize: 12pt
geometry: margin=2.5cm
bibliography: references.bib
header-includes:
  - \usepackage{fancyhdr}  # Custom headers/footers
  - \pagestyle{fancy}       # Enable fancy headers/footers
  - \fancyhf{}              # Clear default headers
  - \lfoot{`r Sys.Date()`}    # Left footer: Author name(s)
  - \rfoot{Page \thepage}   # Right footer: Page number
  - \cfoot{Nathann Morand, Felipe Ramirez, Gautier Demierre} # Center footer
  - \thispagestyle{fancy}   # Ensure first page gets the footer too
---

# A1-10, ANOVA Potato

## Introduction
This study investigates the cooking quality of Oregon-grown Russet potatoes. Specifically, it examines how different growing areas, storage conditions, and cooking methods relate to the flavor of the potatoes. The `Flavor score` is modeled as a function of:

- **Growing Area**: Southern Oregon vs. Central Oregon
- **Two-week Holding Temperature**: 75°F vs. 40°F
- **Size**: Large vs. Medium
- **Storage Period**: 0, 2, 4, and 6 months
- **Cooking Method**: Boiling, Steaming, Mashing, Baking at 350°F, Baking at 450°F


```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Load required libraries
library(knitr)
library(kableExtra)
library(dplyr)
library(tibble)
library(car)
library(ggplot2)
library(broom)  # For clean ANOVA tables
library(glue)
library(gt)

# Load data
potato <- read.table("potato.dat", header = FALSE)
colnames(potato) <- c("Area", "Temp", "Size", "Storage", "Cooking", "Texture", "Flavor",  "Moistness")
#str(potato)

#sum(is.na(potato))

# Convert categorical variables to factors with meaningful labels
potato$Area <- factor(potato$Area, levels = c(1, 2), labels = c("Southern Oregon", "Central Oregon"))
potato$Temp <- factor(potato$Temp, levels = c(1, 2), labels = c("75°F", "40°F"))
potato$Size <- factor(potato$Size, levels = c(1, 2), labels = c("Large", "Medium"))
potato$Storage <- factor(potato$Storage, levels = c(1, 2, 3, 4), labels = c("0 months", "2 months", "4 months", "6 months"))
potato$Cooking <- factor(potato$Cooking, levels = c(1, 2, 3, 4, 5), labels = c("Boil", "Steam", "Mash", "Bake@350°F", "Bake@450°F"))

```

## Exploratory Data Analysis (EDA)
```{r sumarrystat, fig.align='center', fig.width=5, fig.height=4, echo=FALSE, warning=FALSE, message=FALSE, fig.pos='H'}

# Split dataset into categorical and numerical subsets
categorical_vars <- potato[, sapply(potato, is.factor)]
numeric_vars <- potato[, c("Texture", "Flavor", "Moistness")]

# Summary table (same code as avant)
summary_numeric <- numeric_vars %>%
  summarise(
    Min = sapply(., min, na.rm = TRUE),
    `1st Quartile` = sapply(., quantile, probs = 0.25, na.rm = TRUE),
    Median = sapply(., median, na.rm = TRUE),
    Mean = sapply(., mean, na.rm = TRUE),
    `3rd Quartile` = sapply(., quantile, probs = 0.75, na.rm = TRUE),
    Max = sapply(., max, na.rm = TRUE)
  )
summary_num_df <- summary_numeric %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "Statistic")
colnames(summary_num_df) <- c("Statistic", "Texture", "Flavor", "Moistness")
summary_num_df[, 2:4] <- lapply(summary_num_df[, 2:4], function(x) format(round(as.numeric(x), 2), nsmall = 2))

summary_categorical <- lapply(categorical_vars, summary)
summary_cat_df <- bind_rows(lapply(summary_categorical, function(x) as.data.frame(t(x))), .id = "Variable")
summary_cat_df[is.na(summary_cat_df)] <- ""

kable(summary_cat_df, caption = "Summary Statistics (Categorical Variables)", longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, angle = 90)

```

```{r sumarrystatnumerical, fig.align='center', fig.width=7, fig.height=4, echo=FALSE, warning=FALSE, message=FALSE, fig.pos='H'}

library(dplyr)
library(knitr)
library(kableExtra)

library(dplyr)
library(knitr)
library(kableExtra)

get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Compute statistics manually
summary_stats <- tibble(
  Statistic = c("Mean", "Median", "Mode", "SD", "Min", "Max"),
  Texture = c(mean(potato$Texture), median(potato$Texture), get_mode(potato$Texture),
              sd(potato$Texture), min(potato$Texture), max(potato$Texture)),
  Flavor = c(mean(potato$Flavor), median(potato$Flavor), get_mode(potato$Flavor),
             sd(potato$Flavor), min(potato$Flavor), max(potato$Flavor)),
  Moistness = c(mean(potato$Moistness), median(potato$Moistness), get_mode(potato$Moistness),
                sd(potato$Moistness), min(potato$Moistness), max(potato$Moistness))
)

# Round for display
summary_stats[, 2:4] <- round(summary_stats[, 2:4], 2)

# Display table
kable(summary_stats, caption = "Summary Statistics of Numerical Variables") %>%
  kable_styling(latex_options = c("hold_position"))

```

```{r histogramnumerical, fig.cap='Histograms of all numerical variables.', fig.align='center', fig.width=7, fig.height=4.5, echo=FALSE, warning=FALSE, message=FALSE, fig.pos='H'}
# Reshape the data into long format for ggplot
library(tidyr)

potato_long <- potato %>%
  pivot_longer(cols = c(Texture, Flavor, Moistness), names_to = "Variable", values_to = "Value")

# Plot histograms using ggplot2
ggplot(potato_long, aes(x = Value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black", alpha = 0.8) +
  facet_wrap(~ Variable, scales = "free", ncol = 1) +
  theme_minimal() +
  labs(title = "Histograms of Texture, Flavor, and Moistness",
       x = "Score", y = "Frequency")

```

```{r scatterplot, fig.cap='scatterplot and correllation score between every pair of numerical variable', fig.align='center', fig.width=7, fig.height=3, echo=FALSE, warning=FALSE, message=FALSE, fig.pos='H'}

library(GGally)

# Select only numerical variables
numeric_vars <- potato[, c("Texture", "Flavor", "Moistness")]

# Create pairwise scatter plot matrix
ggpairs(numeric_vars,
        lower = list(continuous = wrap("points", alpha = 0.6, size = 1.5)),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
        upper = list(continuous = wrap("cor", size = 4))) +
  theme_minimal()
  
```

\newpage 
Overall, the class are well balanced, the Moistness and Texture score are strongly correlated. Their distribution is nicely behaved but the Flavor score has some hole.

## Model Fitting

The general form of our ANOVA model is:

\[
\begin{aligned} 
Y &= \beta_0 \\  
&+ \beta_1 \cdot \text{Area}_{\text{Central Oregon}} \\  
&+ \beta_2 \cdot \text{Temp}_{40^{\circ}F} \\  
&+ \beta_3 \cdot \text{Size}_{\text{Medium}} \\  
&+ \beta_4 \cdot \text{Storage}_{\text{2 months}} \\  
&+ \beta_5 \cdot \text{Storage}_{\text{4 months}} \\  
&+ \beta_6 \cdot \text{Storage}_{\text{6 months}} \\  
&+ \beta_7 \cdot \text{Cooking}_{\text{Steam}} \\  
&+ \beta_8 \cdot \text{Cooking}_{\text{Mash}} \\  
&+ \beta_9 \cdot \text{Cooking}_{\text{Bake@350}^{\circ}F} \\  
&+ \beta_{10} \cdot \text{Cooking}_{\text{Bake@450}^{\circ}F} \\
&+ \varepsilon
\end{aligned}
\]

where:  

- $Y$ is the predicted **Flavor Score**.  
- $\beta_i$ are the coefficients associated with the explanatory variables.
- $\text{Area}_{\text{Central Oregon}}$: effect of being grown in **Central Oregon** compared to **Southern Oregon**.
- $\text{Temp}_{40^{\circ}F}$: effect of storing the potato at **40°F** compared to **75°F**.
- $\text{Size}_{\text{Medium}}$: effect of using **Medium-sized** potatoes compared to **Large** ones.
- $\text{Storage}_{\text{2 months}}$: effect of storing the potato for **2 months** compared to **0 months**.
- $\text{Storage}_{\text{4 months}}$: effect of storing the potato for **4 months** compared to **0 months**.
- $\text{Storage}_{\text{6 months}}$: effect of storing the potato for **6 months** compared to **0 months**.
- $\text{Cooking}_{\text{Steam}}$: effect of **Steaming** compared to **Boiling**.
- $\text{Cooking}_{\text{Mash}}$: effect of **Mashing** compared to **Boiling**.
- $\text{Cooking}_{\text{Bake@350}^{\circ}F}$: effect of **Baking at 350°F** compared to **Boiling**.
- $\text{Cooking}_{\text{Bake@450}^{\circ}F}$: effect of **Baking at 450°F** compared to **Boiling**.
- $\varepsilon$: the **residual error**, capturing variation not explained by the model.

The model is fitted using the least square method. 
The model do not take interraction into account as it would likely lead to a poor fit given the limited number of observation.
The model is selected by minimizing the Akaike Information Criterion (AIC). This work in a backward manner where parameters are itterratively eliminated to reach a good trade of between model strength and simplicity. 


```{r modelfittingselect, echo=FALSE, , message=FALSE, warning=FALSE}
# Fit ANOVA model
# Load required package
library(MASS)

# Fit the full model (all main effects)
anova_model <- lm(Flavor ~ Area + Temp + Size + Storage + Cooking, data = potato)

# Perform stepwise model selection using AIC
model_step <- stepAIC(anova_model, direction = "both", trace = FALSE)

# Create a comparison table of AIC values
aic_comparison <- data.frame(
  Model = c("Original (Full)", "Final (StepAIC)"),
  AIC = c(AIC(anova_model), AIC(model_step))
)

# Print AIC comparison table
library(knitr)
library(kableExtra)

kable(aic_comparison, caption = "Comparison of AIC Before and After Model Selection") %>%
  kable_styling(latex_options = c("hold_position"))


```

## Model assessment
For the ANOVA model to be valid, the following assumptions must hold:  

Errors have zero mean: On average, the residuals should be centered around zero.

Errors are homoscedastic: The residuals should have constant variance across all levels of the predictors.

Errors are uncorrelated: There should be no systematic patterns or autocorrelation among the residuals.

Errors are normally distributed: The residuals should approximately follow a normal distribution.

```{r, fig.cap='QQ plot of residual and residuals vs fitted', fig.align='center', fig.width=7, fig.height=4, echo=FALSE, warning=FALSE, message=FALSE, fig.pos='H'}
par(mfrow = c(1, 2))  # Arrange plots in one row

# Q-Q Normal Plot of Residuals
qqnorm(residuals(anova_model), main = "Q-Q Plot of Residuals")
qqline(residuals(anova_model), col = "red")

# Residuals vs Fitted Values
plot(fitted(anova_model), residuals(anova_model),
     main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

```

To assess whether the ANOVA assumptions hold, we examine residual diagnostics.

The Q–Q plot shows that the residuals follow the red line closely, especially in the center. That means the residuals are approximately normally distributed, which supports the normality assumption.

In the residuals vs fitted plot, the points are scattered randomly around zero with no clear pattern. That suggests the variance is roughly constant (homoscedasticity). There's no obvious curve, so the model seems appropriate.

Overall, the residual diagnostics support the assumptions of the ANOVA model.

## Final model

The estimated model with numeric values is:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Function to escape LaTeX special characters
sanitize_latex <- function(name) {
  name <- gsub("°", "^{\\circ}", name, fixed = TRUE)  # Convert degree symbol
  return(name)
}


# Extract coefficients
coeffs <- coef(model_step)
coeff_values <- signif(coeffs, 2)  # Round to 2 significant digits
variable_names <- names(coeffs)[-1]  # Exclude intercept
variable_names <- sapply(variable_names, sanitize_latex)  # Apply sanitization

# Start LaTeX equation
equation_text <- "\\begin{aligned} \\hat{Y} &= "  

# Add intercept (β₀)
equation_text <- paste0(equation_text, signif(coeffs[1], 2), " \\\\ ")  

# Loop through remaining coefficients
for (i in seq_along(variable_names)) {
  sign <- ifelse(coeff_values[i + 1] >= 0, "&+ ", "&- ")  # Align sign
  equation_text <- paste0(equation_text, sign, abs(coeff_values[i + 1]), " \\cdot ", variable_names[i], " \\\\ ")
}

# End equation (no epsilon term)
equation_text <- paste0(equation_text, "\\end{aligned}")

```
$$
`r equation_text`
$$


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Residuals from the final model
res <- residuals(model_step)

# Compute error metrics
MAE <- mean(abs(res))
RMSE <- sqrt(mean(res^2))
SD_resid <- sd(res)  # Standard deviation of residuals (often called sigma)

# Display results
error_metrics <- data.frame(
  Metric = c("Mean Error (should be ~0)", "Mean Absolute Error", "Root Mean Squared Error", "Std. Dev. of Residuals"),
  Value = round(c(mean(res), MAE, RMSE, SD_resid), 3)
)

kable(error_metrics, caption = "Error Metrics for Final ANOVA Model") %>%
  kable_styling(latex_options = c("hold_position"))

```

\newpage 

## Conclusion

This study analyzed the factors linked to the **flavor score** of Oregon-grown Russet potatoes using an **ANOVA model**. The key variables considered included **growing area, storage conditions, size and cooking methods**.  
At first an exploratory data analysis was conducted. It allowed to assess that the class where well balanced and that the Flavor score where well distributed. The model was then defined and fitted using Least Square. It was refined by removing parameters to minimize the AIC. Finally the model was assessed to ensure the required assumption held and was written numerically.

### **Main Findings**
1. **Significant Factors**
   - **Holding temperature (75°F vs. 40°F)** had the strongest correlation with flavor, with temperatures positively linked to flavor scores.  
   - **Storage duration** is linked to flavor, with 2 to 6 month storage having a stronger positive corellation with flavor compared to fresh potatoes.  
   - **Cooking method** is also linked to flavor, with high-temperature baking (450°F) and steam cooking being negatively corellated with the flavor scores, while baking at 350°F** and mashed cooking was positively correlated.  
   - **Growing in Central Oregon** was also linked positively with score.  

2. **Insignificant Factor**
   - **Potato size (Large vs. Medium)** was not corellated and was thud dropped when refinning the model.

### **Model Fit & Assumption Verification**
- The **residual analysis** (Q-Q plot and residuals vs. fitted values) indicated that the model mets normality, homoscedasticity assumptions along with the erros being normally distributed and its mean zero. The root mean squared error was 0.22
- The estimated residual standard deviation was **0.221**, confirming a **good model fit**.  


# References
\renewcommand\refname{}
\renewcommand\bibname{}
\nocite{*}

